\section{Background and Previous System} \label{sec:background-and-previous-system}
This chapter establishes the baseline for the thesis.
It first introduces a browser-based Linux command-line practice environment and summarizes prior development efforts, starting from the initial prototype and continuing with the later bachelor's thesis implementation.
It then outlines the architecture and Azure deployment of the previous system that serves as the starting point for this master's thesis.
Finally, the chapter identifies the main limitations and outlines the improvement opportunities that guide the design and evaluation in later chapters.

\subsection{Motivation and Prior Work} \label{subsec:motivation-and-prior-work}
Linux command-line knowledge remains an important skill in many computing roles, particularly in system administration and automation.
Consequently, hands-on command-line practice is a relevant educational goal.
Earlier development work at the University of Tartu explored how such practice can be supported using a browser-based platform, beginning with an initial prototype and progressing to a more stable cloud-hosted implementation.

\subsubsection{The Importance of the Linux Command Line} \label{subsubsec:the-importance-of-the-linux-command-line}
Proficiency in command-line operations is a valuable skill due to the widespread use of Linux-based systems in modern computing infrastructure, which are often operated and maintained through terminal-centric workflows.
An observable indicator of Linux's prevalence in production computing environments is provided by high-performance computing.
According to TOP500 statistics, Linux is the operating system family used by all TOP500-ranked supercomputers~\cite{noauthor_operating_nodate}.
In such environments, administration, troubleshooting, and automation typically depend on command-line tools and remote terminal access instead of graphical interfaces.

This view aligns with academic perspectives on skills and learning.
Švábenský et al.~\cite{svabensky_toolset_2021} characterize the Linux command-line interface as common both in higher education computing environments and in software development practices.
They also advocate for studying command-line interactions to enhance comprehension and facilitate hands-on learning.
Likewise, Bailey and Zilles~\cite{bailey_uassign_2019} treat terminal skills as a topic worth teaching and assessing through authentic terminal actions, and they present an approach for interactive, auto-graded terminal assignments.

At the University of Tartu, Linux and its command line are treated as core topics in systems-oriented teaching.
For example, Linux command-line work plays a central role in the ``Operating Systems'' and ``Computer Security'' courses.
This instructional need motivated the development of a browser-based learning environment in two main iterations.
The first was a proof-of-concept prototype described by Halapuu~\cite{halapuu_eestikeelse_nodate}, and the second was a cloud-hosted implementation presented in Eistre's bachelor's thesis~\cite{eistre_veebipohise_nodate}.
These two iterations are summarized in the following subchapters.

\subsubsection{Initial Proof-of-Concept} \label{subsubsec:initial-proof-of-concept}
Halapuu~\cite{halapuu_eestikeelse_nodate} presents the first iteration of the browser-based Linux command-line learning environment.
In Halapuu's thesis, the prototype provides students with a web terminal connected to an isolated Linux environment, enabling interactive command execution and task-based learning within the browser.
The prototype also includes an automatic task-checking approach based on observable outcomes such as command output and filesystem changes, which allowed students to receive immediate feedback and progress through tasks.

Halapuu~\cite{halapuu_eestikeelse_nodate} further reports that the proof-of-concept application was deployed inside the University's internal network and run on the supervisor's university laptop.
The platform was tested in the ``Computer Security'' course context with over one hundred participants.
Based on the reported evaluation, he concludes that the overall approach was viable for course use, while also identifying improvement opportunities related to stability, security, and maintainability.

\subsubsection{Cloud-Hosted Implementation} \label{subsubsec:cloud-hosted-implementation}
After the initial proof-of-concept had been used in the supervisor's courses for two years, Eistre~\cite{eistre_veebipohise_nodate} advanced the system into a cloud-hosted platform designed for sustained course use.
In Eistre's thesis, the work addresses limitations identified during previous use and sets goals such as improving stability and maintainability, supporting both Estonian and English, and deploying the system in a more robust cloud environment.

Eistre~\cite{eistre_veebipohise_nodate} reports that the resulting environment was evaluated in the context of the ``Computer Security'' course in February 2024.
During this period, the platform was used by 382 participants.
He concludes that the overall outcome was successful, while also noting minor issues observed during the session, which informed further improvements for continued production use.

The following section outlines the architecture and cloud deployment of the baseline system, which serves as the starting point for the work presented in this thesis.
