%! suppress = UnresolvedReference
% TODO check footnotes
\section{Revised Architecture, Implementation, and Development Process} \label{sec:revised-architecture-implementation-and-development-process}
This chapter outlines the modernized architecture and implementation of the Linux command-line learning environment.
Building upon the baseline system described in Chapter 2, the new design prioritizes modularity, improved maintainability, and cost-efficient operation.
The following sections detail the revised system architecture, key technological choices, and the implementation of essential components.
Collectively, these modifications seek to mitigate the shortcomings of the prior version while maintaining the platform's fundamental functionality.

\subsection{Development Process and AI-assisted Refactoring Workflow} \label{subsec:development-process-and-ai-assisted-refactoring-workflow}
The implementation work in this thesis required coordinated refactoring across multiple layers of the baseline system while keeping the codebase operational throughout the change.
To speed up this process, the author used LLM-based (large language model) coding assistants as interactive tools for refactor planning, scaffolding, and iterative debugging.
However, final responsibility for design decisions, integration, and verification remained with the author.

\subsubsection{AI-assisted Development} \label{subsubsec:ai-assisted-development}
Recent software engineering literature describes LLM-assisted development as a shift toward more conversational, intent-driven workflows, where developers specify goals and constraints in natural language and use AI tools to scaffold or generate implementation artifacts.
Beaulieu et al.~\cite{beaulieu_vibe_2025} frame this as a move away from syntax-centric programming toward more dialog-based and intent-driven collaboration with AI tools, and they connect such tools to reduced repetitive manual effort.
Similarly, De Silva et al.~\cite{de_silva_generative_2025} describe vibe coding as a conversational natural-language process for prototype development, where functionality is gradually developed through collaboration with an AI coding assistant and refined through successive prompts.

At the same time, the literature emphasizes that these productivity gains require careful oversight.
De Silva et al.~\cite{de_silva_generative_2025} warn that conversational workflows can encourage acceptance of generated code with limited understanding or review, which increases the risk of hidden defects and non-functional risks such as security, performance, and maintainability problems.
They recommend human review, critical evaluation of AI outputs, and hybrid workflows that combine AI assistance with conventional engineering practice.
Malamas et al.~\cite{malamas_toward_2025} further highlight that while conversational interaction can speed up prototyping, generated code may still be erroneous or insufficiently structured, requiring developer control and review for successful integration.

In this thesis, AI-assisted coding was therefore treated as a development support method rather than as an autonomous implementation process.
The tools were used to speed up planning, draft boilerplate, suggest refactor steps, and help interpret concrete error messages, but architectural decisions, code integration, and final acceptance of changes remained the responsibility of the author.

\subsubsection{Refactoring Workflow} \label{subsubsec:refactoring-workflow}
% TODO add references for the concrete tools and models used, or move these details to an appendix after the next Zotero export.
AI assistance was primarily used through OpenCode, a command-line coding application, together with GitHub Copilot, via GitHub Education, for model access.
The primary models used were Claude Sonnet 4.5 and GPT-5.2-Codex, selected based on availability and suitability for coding and debugging tasks.

Most of the significant changes, such as the container provisioning abstraction and the localisation redesign, followed a consistent workflow:

\begin{enumerate}
\item Scope and constraint definition: define a concrete implementation goal together with technical constraints and acceptance conditions;
\item Prompt-driven planning and design exploration: ask the assistant to analyze the existing code and propose a development plan;
\item Plan iteration: refine the proposed approach through discussion until the intended change scope is clear;
\item Scaffold and draft code changes: generate boilerplate and first-pass implementations to reduce repetitive manual work;
\item Review and integration: inspect, simplify, and adapt the generated output before integrating it into the codebase;
\item Verification and iteration: validate the result through type checking, linting, and manual testing, then use concrete failures as input for follow-up fixes.
\end{enumerate}

% TODO
One example prompt used for refactoring is shown below:

PROMPT PLACEHOLDER - use code block or something similar

Given the risk of integrating incorrect or overly complex generated changes, explicit validation steps were applied.
Large refactors were divided into smaller, reversible steps, and changes were only accepted after passing type checks, linting, and manual verification.
Additionally, security- and cost-sensitive areas received stricter review.
These measures align with De Silva et al.~\cite{de_silva_generative_2025} and Malamas et al.~\cite{malamas_toward_2025}, who emphasize that LLM-driven iteration still requires conventional engineering rigor to prevent issues related to correctness and maintainability.

\subsection{Architecture Overview} \label{subsec:architecture-overview}
The revised system architecture separates the interactive learning workflow from infrastructure-specific concerns and periodic maintenance.
This design preserves the user-facing workflow while allowing the execution backend and operational mechanisms to differ across deployments.

% TODO
PLACEHOLDER FIGURE 5. Revised architecture.

Figure 5 groups the system into four conceptual components.
The first is the browser, responsible for rendering the web interface and hosting the interactive terminal.
The second is the main application environment.
It contains the primary Nuxt server and the scheduled maintenance jobs.
In this context, the Nuxt server provides both the Vue.js user interface and the API surface used by the frontend.
The third component is the database, which retains the persistent application state required for continuity and administration, including user data, progress, and educational content.
The fourth component is the container execution environment, which provides isolated Linux sessions for users during terminal-based exercises.

A significant architectural change compared to the baseline system is the separation of maintenance tasks from the primary application into dedicated runnable functions deployed as scheduled jobs.
As illustrated in Figure 5, these are represented by the provisioner cleanup cronjob and the database cleanup cronjob.
The provisioner cleanup job enforces lifecycle policies for user container environments, such as removing expired sessions, while the database cleanup job removes expired verification tokens and inactive user records in accordance with retention policies.
This separation prevents the main application process from running indefinitely solely to perform periodic maintenance tasks.
It also enables deployments where the main application can be scaled down during periods of inactivity when it is cost-effective, while scheduled maintenance continues to execute and preserve system consistency.

During an active terminal session, the browser establishes a persistent connection to the backend API to support interactive input and output.
The backend then bridges this traffic to the user's Linux session in the container execution environment.
The details of how the user session is provisioned and how lifecycle rules are implemented are covered in later sections.

This overview remains provider-agnostic and focuses on the conceptual decomposition of the system.
The following sections describe the modularization of the codebase, the provisioning abstraction and lifecycle management, the authentication and authorization design, and the improved data model and localisation support.

\subsection{Modularization, Modernization, and Codebase Structure} \label{subsec:modularization-modernization-and-codebase-structure}
The revised implementation restructures the codebase around explicit modular boundaries.
The primary objective is to reduce coupling between unrelated concerns, thereby ensuring that changes remain local, more comprehensible, and less prone to introducing regressions throughout the system.
This approach aligns with the fundamental principles of modularity.
Parnas~\cite{parnas_criteria_1972} argues that the effectiveness of modularization depends on the system's decomposition, highlighting that good decomposition enhances flexibility and comprehensibility by concealing design decisions within module boundaries.
Recent work on modular monolith architectures reports similar maintainability and reduced operational overhead benefits in modern cloud environments~\cite{al-qoran_modular_2025}.

The new code repository layout is organized as a monorepo with two top-level directories, /apps and /packages.
The /apps directory contains runnable units that can be deployed independently, including the main Nuxt web application, which hosts both the Vue.js user interface as well as the API, and the two scheduled maintenance jobs mentioned in the previous subchapter.
The /packages directory contains reusable service modules that are imported by the runnable apps.
These modules encapsulate specific functionality behind stable interfaces, such as provider-agnostic container provisioning, session handling, and configuration.
This allows the runnable apps to remain comparatively thin orchestration layers, depending on the interfaces rather than the internal details of the modules.

This restructure also supports the architectural goal of decoupling maintenance from the main interactive learning workflow.
Both scheduled jobs in the /apps directory reuse the same core modules in /packages as the main Nuxt application, but run as separate executables on a schedule.
This avoids keeping the main application running indefinitely solely to perform cleanup and enables deployments where the interactive application can be scaled down during idle periods while maintenance continues to enforce lifecycle and retention policies.

The modernization effort also included updating core frameworks and supporting dependencies to currently supported stable versions where feasible.
One example was the migration of the main application from Nuxt 3 to Nuxt 4, which addressed the outdated dependency baseline identified in Section 2.3.3.
In this way, dependency updates were treated as part of the broader codebase modernization effort rather than as isolated package-management tasks.

Finally, clearer modular boundaries improve validation, readability, and maintainability by enabling more localized reasoning and testing.
Structuring the codebase into reusable packages and independently runnable apps makes it easier to extend in the future without destabilizing the overall workflow of the application.
The next section builds on this structure by describing the provisioning abstraction and the lifecycle management approach implemented behind the provisioning interface.

\subsection{Container Provisioning and Lifecycle Management} \label{subsec:container-provisioning-and-lifecycle-management}
The updated system introduces a provider-agnostic provisioning layer responsible for the creation, maintenance, and deletion of the user-specific Linux containers required for interactive terminal sessions.
This is a central change from the baseline, as it encapsulates provider-specific runtime logic behind a stable API, allowing the same user-facing workflow to operate across different execution backends.

\subsubsection{Provisioner Interface and Shared Abstraction} \label{subsubsec:provisioner-interface-and-shared-abstraction}
To decouple the core application from provider-specific details, the system defines an abstract provisioning interface that all backends must implement.
Upon application startup, a factory function evaluates the environment configuration and instantiates the appropriate provisioner implementation.
This approach ensures that the application relies solely on a stable API, while provider-specific logic is contained within the implementation classes.

At a high level, the provisioner supports four operations.
It can list currently active user containers, ensure a user container exists and create one if it is not present, update the container expiration time while the user remains active, and delete the container once it is no longer required.
The provisioner interface is shown in the code snippet below.

% TODO
TODO PROVISIONER INTERFACE SNIPPET
```
export interface Provisioner {
  /**
  * List user IDs and expiration times for all containers.
  * @returns List of user IDs and expiration times
  */
  listContainers: () => Promise<ContainerInfo[]>;

  /**
  * Ensure a container exists for the given user ID.
  * If it does not exist, create it.
  * @param userId Unique identifier for the user
  */
  ensureContainerExists: (userId: string) => Promise<ConnectionInfo>;

  /**
  * Update the expiration time of an existing container.
  * @param userId Unique identifier for the user
  */
  updateContainerExpiration: (userId: string) => Promise<void>;

  /**
  * Delete a container and all associated resources.
  * @param userId Unique identifier for the user
  */
  deleteContainer: (userId: string) => Promise<void>;
}
```

The provisioning interface is implemented by classes specific to each provider, while shared operational concerns are centralized in a common base class, AbstractProvisioner.
In this base class, each provisioning operation is wrapped with a concurrency limiter to cap simultaneous operations against provider APIs, a per-user lock to serialize operations per user, and retry logic to handle transient failures.
For concurrency limiting and retry handling, the p-limit and p-retry packages are used.
The following code snippet demonstrates this wrapping for one of the interface functions, illustrating how the public method enforces shared guarantees while delegating provider-specific work to the underlying implementation method.

% TODO
TODO ABSTRACT WRAPPER SNIPPET
```
export abstract class AbstractProvisioner implements Provisioner {
  /**
  * @inheritDoc
  * @param userId Unique identifier for the user
  */
  async ensureContainerExists(userId: string): Promise<ConnectionInfo> {
    return this.withConcurrencyLimit(() =>
    this.withLock(userId, () =>
    this.withRetry(() => this.ensureContainerExistsImpl(userId))),
    );
  }

  protected abstract ensureContainerExistsImpl(userId: string): Promise<ConnectionInfo>;
}
```

In practice, these wrappers are important under classroom-scale usage patterns and bursty traffic, such as page refreshes and repeated session starts.
Per-user serialization prevents race conditions where concurrent requests attempt to create or delete the same user's container.
Concurrency limiting reduces the likelihood of overwhelming provider APIs during spikes, and retry logic improves robustness against temporary network failures or throttling.
Additionally, the base class provides shared utilities for lifecycle management, including computing expiration timestamps, applying consistent metadata tags or labels, and generating fresh SSH key pairs for new containers, ensuring consistent behaviour across providers.

\subsubsection{Lifecycle Model, Cleanup, and Provider Implementations} \label{subsubsec:lifecycle-model-cleanup-and-provider-implementations}
The provisioner abstraction presents a provider-agnostic API. For example, listContainers() returns a list of user identifiers and expiration timestamps for all managed containers, regardless of whether the underlying runtime is Kubernetes or another backend.
Likewise, ensureContainerExists(userId) guarantees that the system can obtain connection information for a ready interactive container, even though the internal handling differs per provider.
Provider implementations therefore encapsulate the relevant state transitions, such as handling the absence of a container by creating one, waiting or polling while a container is starting, and refreshing expiration metadata when a container is already running.

Currently, three container runtime providers are supported.
These are Kubernetes, Azure, and AWS. All implementations follow the same lifecycle model by attaching common metadata to managed resources, such as the user identifier and expiration time, enabling consistent enumeration and reclamation of inactive containers.
The deployment setup and provider-specific configuration are described in Chapter 4.

Lifecycle management is treated as a first-class operational concern to prevent resource leakage and control costs.
User containers are intentionally time-bound and carry an inactivity-based expiration timestamp that is extended during activity via updateContainerExpiration().
Enforcement of expiration is separated from the main web application via the scheduled provisioner cleanup job described in Section 3.2. At a high level, the cleanup job periodically lists managed containers using listContainers(), identifies those whose expiration time is in the past, and deletes them through deleteContainer(userId).
This separation ensures that expired containers are reclaimed even if the main application is scaled down during idle periods, aligning the operational footprint more closely with actual usage.
A simplified version of the cleanup flow (excluding logging) is shown below.

% TODO
PLACEHOLDER FOR PROVISIONER CLEANUP SNIPPET
```
try {
  // List all containers
  const containers = await provisioner.listContainers();

  // Filter expired containers
  const now = new Date();
  const expired = containers.filter(container => container.expiresAt <= now);

  // Delete expired containers
  const results = await Promise.allSettled(
    expired.map(container => provisioner.deleteContainer(container.userId)),
  );
}
catch (error) {
  process.exit(1);
}
```

Although this subsection focuses on the provisioner module because it is the most complex, the same interface-driven design pattern is applied across other internal packages.
For example, the mailer package is structured around a provider boundary, including implementations for both standard SMTP (Simple Mail Transfer Protocol) and Azure Communication Services, the session module encapsulates the WebSocket and SSH bridging responsibilities behind a stable API, and the evaluator package isolates output buffering and regex-based validation logic.
Even where only a single implementation exists today, these boundaries keep responsibilities explicit and reduce the likelihood that infrastructure or operational concerns leak into the user-facing application.

\subsection{Authentication and authorization} \label{subsec:authentication-and-authorization}
In the authentication layer, the updated system replaces the Lucia package with an actively maintained solution that supports session management and single sign-on (SSO) based on OAuth 2.0 and OpenID Connect (OIDC). This is a central change from the baseline, as it removes dependency on a deprecated library discussed in Section 2.3.2 and enables standards-based identity integration for long-term maintainability and secure access control.

\subsubsection{Authentication Design and Implementation} \label{subsubsec:authentication-design-and-implementation}
Authentication is essential for this learning platform to identify users and ensure continuity of sessions for individual Linux container environments.
User accounts enable the system to associate and store task results for each student and to limit access to authorized users.
This approach preserves progress across sessions and reduces the risk of misuse or excessive load by requiring authenticated access before provisioning cost-incurring resources.

Given Lucia's deprecation, continuing to rely on it was not suitable for a production system.
As an alternative, the updated system adopts Better Auth, a framework-agnostic authentication library that provides built-in support for common workflows and simplifies long-term maintenance.
Better Auth supports email and password login, session management, role-based authorization, and OAuth 2.0 / OIDC integration for SSO. In addition, the updated system implements an email domain allowlist to restrict registration and login to approved domains, for example, ut.ee.
This helps ensure the platform is used only by the intended audience.

Internally, Better Auth reduces the amount of custom authentication code compared to Lucia by providing standardized handlers and integrations for typical authentication flows.
This decreases the likelihood of implementation errors and makes the authentication layer easier to evolve alongside the rest of the stack.
Overall, the new design preserves the baseline requirements for verification and sessions while improving maintainability and enabling SSO-based onboarding.

\subsubsection{OIDC / OAuth2} \label{subsubsec:oidc-/-oauth2}
OpenID Connect (OIDC) is an authentication protocol built on top of the OAuth 2.0 framework.
Fett et al.~\cite{fett_comprehensive_2016} describe OAuth 2.0 as a widely deployed authorization framework and as the foundation for OpenID Connect.
Fett et al.~\cite{fett_web_2017} further explain that OIDC allows a relying party to delegate authentication to an identity provider and to verify the result using a cryptographically signed id token.

At a high level, the platform uses the OIDC authorization code flow.
As described by Fett et al.~\cite{fett_web_2017}, the user is redirected from the application to the identity provider, authenticates there, and is then redirected back with an authorization code.
They further explain that the relying party exchanges this code for tokens, validates the returned id token, and then establishes a session for the authenticated user.

In the current implementation, this flow can also use PKCE (Proof Key for Code Exchange), which adds an additional verification step during token exchange and improves resistance to authorization code interception.
The web application currently supports both Microsoft Entra ID, which was used during testing, and Keycloak, which is used for university integration, as OIDC providers through environment variable configuration.
After successful token validation, the authenticated external identity is mapped to the platform's internal user and session model, which is then used for progress tracking and authorization decisions.

A standards-based OIDC integration improves interoperability and maintainability compared with provider-specific or custom authentication logic.
Fett et al.~\cite{fett_comprehensive_2016} show that secure OAuth-based systems depend on correct flow handling and validation under recommended best practices, while Fett et al.~\cite{fett_web_2017} provide the corresponding OIDC-specific analysis.
In the context of this platform, OIDC therefore provides a suitable and extensible basis for single sign-on support.

\subsection{Improved localisation and task editor} \label{subsec:improved-localisation-and-task-editor}
The revised system improves both the internal localisation model and the administrative workflow for managing multilingual learning content.
Although the baseline platform already supported multiple interface languages, the updated implementation extends this to the content model itself.
This was achieved by redesigning topics and tasks so that language-independent structure is separated from language-specific text, and by introducing an editor that supports both single-language and dual-language authoring.

\subsubsection{Improved Localisation} \label{subsubsec:improved-localisation}
A significant improvement in the revised implementation is the use of separate translation tables for topics and tasks.
In the updated schema, the topics and tasks tables store the stable logical structure of the educational content, including identifiers, ordering, and task evaluation configurations.
Additionally, the locale-specific textual fields are stored separately in topic_translations and task_translations tables.
Consequently, titles, descriptions, and hints are no longer tied directly to the structural entity, but are instead attached as translations for each locale.

This design improves maintainability and consistency by eliminating the need to duplicate topics or tasks across multiple languages.
It also ensures that the logical components of a task, such as its order, regular expression, or monitored path, are shared across all language variants, thereby minimizing the potential for structural divergence between the Estonian and English versions over time.
Additionally, the uniqueness constraints on entity and locale combinations guarantee that each topic or task is limited to a single translation per language, thereby maintaining a predictable and consistent multilingual content model.

An additional benefit of this structure is that user progress is maintained independently of language.
Task completion is recorded using stable task and topic identifiers rather than specific translation entries.
Consequently, users can switch between supported interface locales without disrupting continuity, while still seeing the same logical exercise sequence in their chosen language.
This enhances the robustness of the localisation model compared to a duplication-based approach, where each language version would function as a separate exercise set.

\subsubsection{Task Editor} \label{subsubsec:task-editor}
To support the revised localisation model, the administrative task editor was redesigned to manage both single-language and dual-language content entries.
Instead of storing a separate configuration flag for editor mode, the implementation determines the editing mode directly from the available topic translations.
The mode is determined from the presence of non-empty topic titles for each locale.
If only one locale contains content, the editor functions in single-language mode for that locale.
Conversely, if both Estonian and English titles are available, the editor transitions to dual-language mode.

This approach keeps the editor logic consistent with the underlying content structure.
The selected editing mode is therefore not an independent state that could drift out of sync with the stored translations, but instead reflects the current multilingual completeness of the topic.
If neither locale is yet populated, the editor falls back to a default locale, allowing content creation to begin in a predictable way.
Figure X presents the editor in single-language mode, where content is entered for only one locale.

% TODO
Figure X. Task editor in single-language mode

In the simpler case, the editor can therefore be used to create or modify a topic or task in only one language.
This is useful when content is intended for a single audience, or when a translation has not yet been prepared.
In this mode, the interface remains compact and reduces unnecessary editing overhead.
When bilingual content is required, the same editor supports entering both language variants within a unified workflow.
Instead of creating separate tasks per language, the editor allows the administrator to manage multiple translations for the same underlying topic or task entity.
This better reflects the revised schema, where one structural task may have several textual representations depending on locale.
Figure Y shows the dual-language mode, where both English and Estonian translations are edited within the same workflow.

% TODO
Figure Y. Task editor in dual-language mode with completeness indicators

The editor also provides lightweight completeness feedback through status indicators shown in the outline and action area.
Tasks or topics with all required values present are marked with an OK indicator, whereas entries with missing mandatory fields are marked as MISSING. This allows administrators to identify incomplete translations or task definitions without opening every item individually, which becomes particularly useful when editing larger bilingual exercise sets.
As a result, the redesign improves editorial workflow by making it easier to maintain parity between language variants while also reducing duplication in content management.

Together, the improvements described in this chapter establish the revised implementation of the platform.
The following chapter builds on this by describing the concrete deployment setups and infrastructure choices used to run the system across different environments.
